Algebraic Graphs with Class

### 2021-05-27 paper/paper.tex
Incorporate feedback from reviews

\usepackage{geometry}
% \geometry{left=3cm,right=3cm,top=3cm,bottom=3cm}

@@

@@

@@

This lead Mokhov to conceive \textit{algebraic
  graphs}~\cite{mokhov2017algebraic}, a sound and complete representation for
graphs. They abstract away from graph representation details and characterize
graphs by a set of axioms. Algebraic graphs have a small safe core of graph
construction primitives and are suitable for implementing graph transformations.
These primitives are represented in the following datatype:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
\hs{Empty} constructs the empty graph, \hs{Vertex} a graph with a single vertex
and no edges. \hs{Overlay} essentially is the union of two graphs and
\hs{Connect} additionally adds edges from all vertices from one graph to all
vertices of the other.

@@

In~\autoref{sec:deep} we will examine the deep embedding (i.e.\
the datatype as opposed to the type class of the preceding sections) regarding
applicability in practice and in verification (\autoref{sec:quotient}).

@@

\section{A type class for algebraic graphs}\label{sec:locale}
This section will cover most parts of the original paper by
Mokhov~\cite{mokhov2017algebraic}. The graph construction primitives of
algebraic graphs will be formally defined and their soundness and completeness
proven. Afterwards we will cover the algebraic structure exhibited by these
primitives when interpreted as graphs. As part of this we will introduce a type class in Haskell which
further abstracts from implementation details. We will also see that different
classes of graphs can be obtained by extending the set of axioms. Instances of
these different classes and a graph transformation library in Haskell are
provided. To conclude this section we will explore the suitability of this type
class for formal verification in Isabelle/HOL.

\subsection{The Core}\label{sec:core}
In the introduction we already briefly saw the four graph construction
primitives. Let us first define the set of (consistent) graphs $\mathcal{G}$
over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
\subseteq V \times V$.
The four graph construction primitives
then are defined as follows. The \textit{empty} graph, denoted $\eps$ is
simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
trivially consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
So overlay is just the union of two graphs, while connect also adds an edge from
each vertex of its first argument to each vertex of its second argument. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
then also $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect G_2 \in
\mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown in the next subsection. Both of these facts were also formally proven in
Isabelle/HOL (see~\autoref{sec:verification}).

As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
related to addition and multiplication, hence, we give connect a higher precedence
than overlay. Let us consider some examples to develop some intuition for the
primitives. They are illustrated in~\autoref{fig:construction}.
\begin{enumerate}[label=(\alph*)]
\item $1 \overlay 2 = (\{1,2\}, \emptyset)$
\item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
\item $1 \connect (2 \overlay 3) = (\{1,2,3\}, \{(1,2), (1,3)\})$
\item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
\end{enumerate}
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-a.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \overlay 2$}
    \centerline{\includegraphics[scale=0.27]{fig/ex-b.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \connect 2$}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-c.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect (2 \overlay 3)$}
  \end{subfigure}
  \hspace{3mm}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-e-new.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect 2 \overlay 2 \connect 3$}
  \end{subfigure}
  \caption{Examples of graph construction. The overlay and connect operations
    are denoted by $\overlay$ and $\connect$, respectively. Illustrations
    from~\cite{mokhov2017algebraic}.\label{fig:construction}} 
\end{figure*}

We already saw how to deeply embed these graph construction primitives in a
datatype. For greater reusability the following type class can be defined.
\begin{minted}{haskell}
  class Graph g where
      type Vertex g
      empty :: g
      vertex :: Vertex g -> g
      overlay :: g -> g -> g
      connect :: g -> g -> g
\end{minted}
The associated type \hs{Vertex g} represents the universe of vertices
$\mathbb{V}$, the remainder of the type class corresponds to $\eps$, single
vertex graphs, $\overlay$ resp.\ $\connect$.

At this point we will only introduce some very basic functions for constructing
graphs. More advanced constructions and transformations will be given
in~\autoref{sec:trafo}. A graph with a single edge is obtained by simply
connecting two vertices:
\begin{minted}{haskell}
  edge :: Graph g => Vertex g -> Vertex g -> g
  edge u v = (vertex u) `connect` (vertex v)
\end{minted}
A graph with only isolated vertices can be constructed from a list of vertices
as follows:
\begin{minted}{haskell}
  vertices :: Graph g => [Vertex g] -> g
  vertices vs = foldr overlay empty . map vertex
\end{minted}
Replacing \hs{overlay} with \hs{connect} in \hs{vertices} leads to a (directed)
clique. Later we will also consider undirected graphs. In that context this
function actually builds a fully connected graph on the given list of vertices.
\begin{minted}{haskell}
  clique :: Graph g => [Vertex g] -> g
  clique vs = foldr connect empty . map vertex
\end{minted}
Following and extending this approach leads to a safe and fully polymorphic
graph transformation library for any concrete graph representation (e.g.\ graphs
in \texttt{containers} or \texttt{fgl}) instantiating this class.

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
The shared identity and the missing annihilating zero for $\connect$
(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
Using decomposition we can actually prove some of the above properties, leading
to the following minimal set of axioms characterizing directed graphs:
\begin{itemize}
\item $(G, \overlay)$ is a commutative semigroup
\item $(G, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
  \connect z \overlay y \connect z$
\item decomposition: $x \connect y \connect z = x \connect y \overlay x
  \connect z \overlay y \connect z$ 
\end{itemize}
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
      {1 \connect 3} \overlay {2 \connect 3}$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
One can easily check that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
They are a direct translation of the fact, that any algebraic graph expression
$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
\textit{canonical form}:
\[
  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
  u \connect v \right)
\]
The completeness of algebraic graphs was also formally proven in Isabelle/HOL
(see~\autoref{sec:verification}).

Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
idempotent overlay operation as $x \preceq y \iff x \overlay y = y$. This is indeed
a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
transitive). In fact this partial order can actually be used as a definition for
the subgraph relation, denoted by $\subseteq$:
\[
  x \subseteq y \coloneqq x \preceq y \iff x \overlay y = y
\]

So far we only considered directed graphs. Undirected graphs can be
obtained from the very same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
introducing this axiom we can prove some of the other axioms of directed graphs.
Hence, undirected graphs are characterized by the following minimal set of
axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) order by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}.

@@

  instance Ord a => Graph (Relation a) where
    type Vertex (Relation a) = a
    empty       = R Set.empty Set.empty
    vertex  x   = R (singleton x) Set.empty
    overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
    connect x y = R (domain x `union` domain y) (relation x `union` relation y `union'
      fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
\end{minted}
Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
of \hs{+} and \hs{*} as shortcuts for \hs{overlay} resp.\ \hs{connect}.

@@

  gmap :: Graph g => (a -> Vertex g) -> GraphFunctor a -> g
  gmap = flip gfor  
\end{minted}
Interestingly this can be used to merge vertices fulfilling some predicate
\hs{p} by mapping them to a single vertex:
\begin{minted}{haskell}
  mergeVertices :: Graph g => (Vertex g -> Bool)
    -> Vertex g -> GraphFunctor (Vertex g) -> g
  mergeVertices p v = gmap $ \u -> if p u then v else u
\end{minted}
In the original paper \hs{gmap} is then used further to construct rather
sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}.

@@

To experiment further we first formalized the available material in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
involves the definition of a \textit{locale}, which for the
purposes of this paper can be thought of as a type class which can be equipped
with axioms. Proving the statements from the paper in Isabelle/HOL works
without problems.

Noschinski formalized fundamental aspects of graph theory in
Isabelle/HOL~\cite{GraphTheory-AFP}. Part of that is essentially the standard
representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with one single
axiom ensuring consistency). Instantiating a locale in Isabelle/HOL (which can be likened to
instantiating a type class in Haskell) requires to prove the axioms attached to
it. Using the definitions from~\autoref{sec:core} for the primitives it is
straightforward to instantiate the algebraic graph locale for the tuple
representation. We also formally prove that the graph construction primitives only
produce consistent graphs. A proof for the completeness of the primitives is
given as well (in the sense that every graph represented by a tuple can be
expressed as an algebraic graph).

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs.
First impressions suggest that it is feasible to base more sophisticated
formalizations of graph theory on algebraic graphs. Although we suspect that
some additional axiom(s) may be required to make for \textit{interesting} graph
theory; as it stands the statement $\eps \neq \hs{vertex u}$ for some $u \in
\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
instantiate the algebraic graph locale with the unit type.

This is an interesting direction for future work for a related reason that the
type class is interesting in Haskell: The type class forms the basis for 
a polymorphic graph transformation library, whereas the locale can be
seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
using the locale axioms becomes available to any graph representation
instantiating the locale. This obviously becomes more interesting the more material is
available in the locale. Haslbeck et al.\ chose a similar approach to implement
and verify Kruskal's algorithm independent of the concrete graph
representation~\cite{Kruskal-AFP}.

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. However, the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
  fold :: Graph g => Graph (Vertex g) -> g
  fold Empty         = empty
  fold (Vertex  x  ) = vertex x
  fold (Overlay x y) = overlay (fold x) (fold y)
  fold (Connect x y) = connect (fold x) (fold y)

  instance Ord a => Eq (Graph a) where
    x == y = fold x == (fold y :: Relation a)
\end{minted}
This approach is also currently implemented in the library available on
\texttt{hackage} (although an adjacency map is used instead of a relation).
However, this is not ideal for multiple reasons. Coupling the datatype to some
other graph representation merely for equality checks is inconvenient at best.
More interestingly algebraic graphs can compactly represent dense graphs.
Consider the graph \hs{clique [1..n]}, which has a linear size representation in
the deep embedding, while it is quadratic when stored as a \hs{Relation}.
Requiring to fully expand graphs to check equality voids this advantage. A
promising approach to alleviate this issue is \textit{modular graph
  decomposition}~\cite{mcconnell2005linear}, as it allows to minimize algebraic
graph expressions. It remains an open question if this decomposition can be
computed efficiently directly on the deep embedding.

>>>>

\usepackage[a4paper]{geometry}
% \geometry{left=3cm,right=3cm,top=3cm,bottom=3cm}

@@

\usepackage{isabelle,isabellesym}

\newcommand*{\term}[1]{{\isaspacing\isastyle \input{#1}}}
\newcommand*{\Term}[1]{{\isaspacing\isastyle #1}}
\newcommand*{\mTerm}[1]{\text{\Term{#1}}}

@@

\let\subsectionautorefname\sectionautorefname
\let\subsubsectionautorefname\sectionautorefname

@@

This led Mokhov to conceive \textit{algebraic
  graphs}~\cite{mokhov2017algebraic}, a sound and complete representation for
graphs. They abstract away from graph representation details and characterize
graphs by a set of axioms. Algebraic graphs have a small safe core of graph
construction primitives and are suitable for implementing graph transformations.
The four construction primitives are the following:
\begin{enumerate}
\item the \emph{empty} graph
\item single \emph{vertex} graphs
\item \emph{overlay}ing two graphs
\item \emph{connect}ing two graphs
\end{enumerate}
Overlay is the union of two graphs while connect also adds
edges from all vertices from one graph to all vertices of the other.

@@

In~\autoref{sec:deep} we will examine the \emph{deep embedding} of the construction
primitives into an inductive datatype and its
applicability in practice and in verification (\ref{sec:quotient}).

@@

\section{A Type Class for Algebraic Graphs}\label{sec:locale}
This section will cover most parts of the original paper by
Mokhov~\cite{mokhov2017algebraic}. The graph construction primitives of
algebraic graphs will be formally defined and their soundness and completeness
proven. Afterwards we will cover the algebraic structure exhibited by these
primitives when interpreted as graphs. As part of this we will introduce a type class in Haskell which
further abstracts from implementation details. We will also see that different
classes of graphs can be obtained by modifying the axioms. Instances of
these different classes and a graph transformation library in Haskell are
presented. To conclude this section we will explore the suitability of this type
class for formal verification in Isabelle/HOL.

\subsection{The Core}\label{sec:core}
Before defining the four graph construction primitives we define the set of
graphs $\mathcal{G}$ over a fixed universe of vertices $\mathbb{V}$. A graph $G
\in \mathcal{G}$ can be represented as a tuple $G=(V,E)$ where $V \subseteq
\mathbb{V}$. A graph is \emph{consistent} if $E \subseteq V \times V$ holds.

The construction primitives
then are defined as follows. The \textit{empty} graph, denoted by $\eps$, is
simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$, denoted by just $v$, are also
trivially consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
So overlay is just the union of two graphs, while connect also adds an edge from
each vertex of its first argument to each vertex of its second argument. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
are consistent, then so are $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect
G_2 \in \mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs, i.e.\ using the construction primitives only yields
consistent graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown shortly (\ref{sec:algebra}). Both of these facts were also formally proven in
Isabelle/HOL (see~\autoref{sec:verification}).

As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
related to addition and multiplication in a semiring, hence, we give connect a
higher precedence
than overlay. Let us consider a few examples to develop some intuition for the
primitives. In these examples the universe of vertices are the natural numbers
and they are illustrated in~\autoref{fig:construction}.
\begin{enumerate}[label=(\alph*)]
\item $1 \overlay 2 = (\{1,2\}, \emptyset)$
\item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
\item $1 \connect (2 \overlay 3) = (\{1,2,3\}, \{(1,2), (1,3)\})$
\item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
\end{enumerate}
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-a.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \overlay 2$}
    \centerline{\includegraphics[scale=0.27]{fig/ex-b.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \connect 2$}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-c.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect (2 \overlay 3)$}
  \end{subfigure}
  \hspace{3mm}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-e-new.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect 2 \overlay 2 \connect 3$}
  \end{subfigure}
  \caption{Examples of graph construction. The overlay and connect operations
    are denoted by $\overlay$ and $\connect$, respectively. Illustrations
    from~\cite{mokhov2017algebraic}.\label{fig:construction}} 
\end{figure*}

We model the graph construction primitives in Haskell with the following type
class:
\begin{minted}{haskell}
  class Graph g where
      type Vertex g
      empty :: g
      vertex :: Vertex g -> g
      overlay :: g -> g -> g
      connect :: g -> g -> g
\end{minted}
The associated type \hs{Vertex g} represents the universe of vertices
$\mathbb{V}$, the remainder of the type class corresponds to $\eps$, single
vertex graphs, $\overlay$ and $\connect$ respectively.

At this point we will only introduce some very basic functions for constructing
graphs. More advanced constructions and transformations will be given
in~\autoref{sec:trafo}. A graph with a single edge is obtained by simply
connecting two vertices:
\begin{minted}{haskell}
  edge :: Graph g => Vertex g -> Vertex g -> g
  edge u v = (vertex u) `connect` (vertex v)
\end{minted}
A graph with only isolated vertices can be constructed from a list of vertices
as follows:
\begin{minted}{haskell}
  vertices :: Graph g => [Vertex g] -> g
  vertices = foldr overlay empty . map vertex
\end{minted}
Replacing \hs{overlay} with \hs{connect} in \hs{vertices} leads to a (directed)
clique. Later we will also consider undirected graphs. In that context this
function actually builds a fully connected graph on the given list of vertices.
\begin{minted}{haskell}
  clique :: Graph g => [Vertex g] -> g
  clique = foldr connect empty . map vertex
\end{minted}
Following and extending this approach leads to a safe and fully polymorphic
graph transformation library for any concrete graph representation (e.g.\ graphs
in \texttt{containers} or \texttt{fgl}) instantiating this class.

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
For a proper semiring an annihilating zero for $\connect$ (i.e.\
$x \connect 0 = 0 = 0 \connect x$) is missing. It is also unusual that the two operations
share their identity $\eps$.

In addition there is the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
      {1 \connect 3} \overlay {2 \connect 3}$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
With decomposition we can prove by equational reasoning that $\overlay$ is
idempotent and that $\eps$ is indeed its identity.
According to Mokhov, the remaining axioms are a minimal set of axioms
\emph{characterizing} algebraic graphs~\cite{mokhov2017algebraic}. From the
original paper it does not become clear what he means by characterization:
Letting $\mathcal{G} = \{1\}$, $\eps = 1$, $1 \overlay 1 = 1$ and
$1\connect 1 = 1$ gives a structure abiding all axioms. This definition of
$\mathcal{G}$ is clearly not (isomorphic to) what we call directed graphs for
any universe of vertices.
We don't know which additional axioms are required to obtain this isomorphism.
However, the statement $\eps \neq vertex\ u$ for any $u \in \mathbb{V}$ does not
follow from the given axioms. Hence, adding an axiom scheme $\forall u \in
\mathbb{V}: \eps \neq vertex\ u$ seems sensible. Additionally $vertex$ probably
needs to be injective (i.e.\ $vertex\ u = vertex\ v \implies u = v$), however, we don't know if this
is sufficient.

One can, however, easily verify that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
They are a direct translation of the fact, that any algebraic graph expression
$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
\textit{canonical form}:
\[
  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
  u \connect v \right)
\]

Mokhov also follows a standard approach of defining a partial order $(\preceq)$ for the
idempotent overlay operation as $x \preceq y \vcentcolon\Leftrightarrow x \overlay y = y$. This is indeed
a partial order under the graph axioms (i.e.\ it is reflexive, antisymmetric and
transitive). In fact this partial order can be used as a definition for
the subgraph relation, denoted by $\subseteq$:
\[
  x \subseteq y \vcentcolon\Leftrightarrow x \preceq y \vcentcolon\Leftrightarrow x \overlay y = y
\]

So far we only considered directed graphs. Undirected graphs can be
obtained from the very same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
introducing this axiom we can prove some of the other axioms of directed graphs.
Hence, according to Mokhov~\cite{mokhov2017algebraic}, undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as used in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) rank by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}. In all of these the same caveats regarding the
term \emph{characterization} from directed graphs apply.

@@

  instance Ord a => Graph (Relation a) where
    type Vertex (Relation a) = a
    empty       = R Set.empty Set.empty
    vertex  x   = R (singleton x) Set.empty
    overlay x y = R (domain x `union` domain y) (relation x `union` relation y)
    connect x y = R (domain x `union` domain y) (relation x `union` relation y `union`
      fromAscList [ (a, b) | a <- elems (domain x), b <- elems (domain y) ])
\end{minted}
Mokhov also defines a \hs{Num} instance for \hs{Relation} which enables the use
of \hs{+} and \hs{*} as shortcuts for \hs{overlay} resp.\ \hs{connect}.

@@

  gmap :: Graph g => (a -> Vertex g) -> GraphFunctor a -> g
  gmap = flip gfor  
\end{minted}
Interestingly this can be used to merge vertices fulfilling some predicate
\hs{p} by mapping them to a single vertex:
\begin{minted}{haskell}
  mergeVertices :: Graph g => (Vertex g -> Bool)
    -> Vertex g -> GraphFunctor (Vertex g) -> g
  mergeVertices p v = gmap $ \u -> if p u then v else u
\end{minted}
In the original paper \hs{gmap} is then used further to construct rather
sophisticated graphs fully polymorphically~\cite{mokhov2017algebraic}, i.e.\
independent from any concrete graph representation.

@@

To experiment further we first formalized the material from the paper in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
the definition of several \textit{locale}s. For the purposes of this paper a
locale be thought of as a type class which can be equipped with axioms. The
following locale is the basis for our formalization.

\vspace{3mm}
\term{terms/locale_primitives}
\vspace{3mm}\newline
Note that it does not have any
axioms, as we reuse it to define locales for different classes of graphs (i.e.\
directed, undirected, etc.). Proving the statements from the paper in Isabelle/HOL works
without problems.

Noschinski formalized fundamental aspects of graph theory in
Isabelle/HOL~\cite{GraphTheory-AFP}. Part of that is essentially the standard
representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with one single
axiom ensuring consistency). Instantiating a locale in Isabelle/HOL (which can be likened to
instantiating a type class in Haskell) requires to prove the axioms attached to
it. Using the definitions from~\autoref{sec:core} for the primitives it is
straightforward to instantiate the algebraic graph locale for the tuple
representation. We also formally prove that the graph construction primitives only
produce consistent graphs. A proof for the completeness of the primitives is
given as well (in the sense that every graph represented by a tuple can be
obtained using the algebraic graph construction primitives).

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs. A
challenge posed by the approach is that a graph is simply a value of some
abstract type \Term{'g}. Thus, we characterize the existence of a vertex or an
edge in a given graph by the subgraph relation (denoted by $\sqsubseteq$ in the
formalization):

\vspace{3mm}
\term{terms/has_elem}
\vspace{3mm}\newline
Equipped with these definitions we can then go on to inductively define when a
list of vertices is a walk in a given graph:

\vspace{3mm}
\term{terms/vwalk_def}
\vspace{3mm}\newline
We define the empty list as well as the list containing only a single vertex of
the graph as walks (\Term{vwalk\_Nil} and \Term{vwalk\_single}). Additionally if
a list with head \Term{v} (note that Haskell's \hs{(:)} is denoted by \Term{(\#)} in
Isabelle) is a walk and $(u,v) \in E$, then prepending \Term{u} also yields a
walk. Now we can prove various statements about walks (e.g.\ appending walks,
splitting walks, etc.). Interestingly, many of these statements follow directly
from the inductive definition of walks, i.e.\ we don't require any of the
algebraic graph axioms to prove them. We only need them when a proof relies on
the fact that $(u,v) \in E$ implies $u \in V$ as well as $v \in V$ (in other
words, the consistency of a graph).

For reachability we took two approaches. First we defined the set of edges of a
graph as follows:

\vspace{3mm}
\term{terms/edge_set_def}
\vspace{3mm}\newline
Reachability then can be defined via the reflexive, transitive closure of the
edge set. The second approach is to say that $v$ is reachable from $u$ if and
only if there is a walk from $u$ to $v$. These are, however, only proof of
concepts as only transitivity of reachability has been shown for both definitions.

Nonetheless, first impressions from these experiments suggest that it is
feasible to base more sophisticated formalizations of graph theory on algebraic
graphs. We did not experiment with adding additional axioms as described
in~\autoref{sec:algebra}. We instantiate the locale with the unit type, where
all primitives simply map to the unit value. This formalizes that the current
set of axioms is insufficient.

Exploring verification using the locale further is an interesting direction for
future work. The Haskell type class forms the basis for a polymorphic graph
transformation library. In the same vein, the locale can be seen as a
\textit{polymorphic} graph formalization
library. Any lemma we can prove using the locale axioms becomes available to any
graph representation instantiating the locale. This obviously becomes more
interesting the more material is available in the locale. Haslbeck et al.\ chose
a similar approach to implement and verify Kruskal's algorithm independent of
the concrete graph representation~\cite{Kruskal-AFP}.
